Tools for identifying unexpectedly low microfossil count sums

### 2019-05-18 Rmd/count_check_MS.Rmd
chironomid2 added

---
title: "Tools for identifying unexpectely low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so. This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets that fail these tests.
journal: "Quaternary Science Reviews"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/journal-of-paleolimnology.csl
---

@@

loadd(bird_singleton_summary)
loadd(testate_summ)
loadd(last_chance_summ)
loadd(pollen_summ, pollenA, pollen1234)
loadd(owens_summ1, owens_summ, owens_est_summ)
loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ)

@@

The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e. few assemblages should have all counts divisable by an integer larger than one, especially if species richness is high.

@@

Analysis of the breeding bird, Owen's Lake, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts should have singletons, and even more should have a greatest common divisor of one. 
This means that these characteristics are potentially useful for identifying data where the count sum is misreported.

@@

It is also possible that the apparently low count sums are a false positive. 
This can only happen in assemblages that have a greatest common divisor greater than one, which is very rare, especially for taxonomically diverse assemblages. If a dataset contains several taxonomically diverse assemblages that have a greatest common divisor of two (or especially if greater than two) doubts should be raised about whether the counts sums are accurately described.

@@

>>>>

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so. This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets fail these tests.
journal: "Quaternary Science Reviews"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/journal-of-paleolimnology.csl
---

@@

loadd(bird_singleton_summary)
loadd(testate_summ)
loadd(last_chance_summ)
loadd(pollen_summ, pollenA, pollen1234)
loadd(owens_summ1, owens_summ, owens_est_summ)
loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)

@@

The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e. few assemblages should have all counts divisible by an integer larger than one, especially if species richness is high.

@@

Analysis of the breeding bird, Owen's Lake diatom, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts have singletons, and an even larger proportion have a greatest common divisor of one. 
Therefore, these characteristics are potentially useful for identifying data where the count sum is misreported.

@@

It is also possible that the apparently low count sums are a false positive. 
However, as taxonomically diverse assemblages with a greatest common divisor greater than one are rare and all the datasets with apparently low count sums contained multiple taxonomically diverse assemblages with apparently low count sums, this is unlikely.

These tests require the assemblage data which are often not archived. 
However, if the count sums are very low it is possible to recognise this from a stratigraphic diagram. 
For example, a chironomid stratigraphy had an assemblage where all five taxa had a relative abundance of about 20%, suggesting a count sum of five, far below the reported fifty chironomids. Another assemblage had six taxa with a relative abundance of just over 10% and a seventh with twice as much, suggesting a count sum of eight chironomids.
On request, the authors sent the data. 
None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.

## Challenges

@@

## Consequences
